---
title: "Colorectal Cancer Detection Using Targeted Serum Metabolic Profiling"
author: ""
date: "08/13/20"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(ComplexHeatmap)
library(ggsignif)


```

## Background

Data set from a study to identify biomarkers using metabolomics for early detection of colorectal cancer

Data set obtained from the Metabolomic Workbench website

Link:https://www.metabolomicsworkbench.org/data/DRCCMetadata.php?Mode=Study&StudyID=ST000284&StudyType=MS&ResultType=1

Collected 234 serum samples from 3 patient groups: 66 CRC, 76 polyp and 92 healthy

Identified 158 metabolites 

Allows for statistical analysis, heatmap generation and data visualization

Focus on metabolites involved in glycolysis since previous literature has shown that changes in glucose metabolism have been observed in CRC cases as well as other types of tumors

***

## Import Data from Excel file

```{r}
library(readxl)
colon_test_excel<- read_excel("CRC raw data for SUR analysis_ updated with patient ID.xlsx")
```
Table with 234 observations (patient samples) and 124 variables (metabolites+extra information)

Table contains only the information from 113 metabolites that were detected reliably and some information regarding gender, cancer status, age etc.

***

## Explore the data

The raw values are very different between each metabolite
```{r}
#Get an idea of the data
summary(colon_test_excel)
```

***

## Explore the data/ Clean data

Identified Class: Tibble Dataframe

Changed missing values to NA

Changed Group names to CRC, Polyp and Healthy

Created subsets to use in statistical analysis later by selecting columns or using the filter command

```{r}
#Identify class
class(colon_test_excel)

#Change missing values to NA
colon_test_excel[colon_test_excel == "-" ] <- NA

#Change Group Names
colon_test_excel$Groups <- recode_factor(colon_test_excel$Groups, `C` = "CRC", `H` = "Healthy", 'P' = "Polyp")

#Create subsets to use for statistical analysis
crc_healthy <- colon_test_excel[1:158, ]
healthy_polyp <- colon_test_excel[67:234,]
crc_polyp <- filter(colon_test_excel, Groups %in% c('CRC', 'Polyp'))
```
***

## Heatmap

Created matrix 

Calculated the z-score using scale() because the raw values are very different between columns

Confirmed that the z-score command worked by checking the mean and standard deviation (mean should be 0 and sd should be 1)

Created a subset and a heatmap for each group

Turned off row clustering so the order of metabolites stays the same and I can compare between groups

Arranged heatmap side by side to observe differences between groups

The heatmap allows us to see that there are differences in metabolite levels between groups as well as within groups

We need to explore the relationships between groups and metabolites more in depth

```{r}
test_matrixall <- as.matrix(colon_test_excel[,3:115], "numeric")
rowlistall <- colon_test_excel[,1]
rowlistall <- unlist(rowlistall)
rownames(test_matrixall)<- rowlistall
zscorematrix <-scale(test_matrixall, center = TRUE, scale = TRUE)

round(apply(zscorematrix [,1:10],2, mean, na.rm = TRUE))
round(apply(zscorematrix [,1:10],2, sd, na.rm = TRUE))

CRC_matrixall <- zscorematrix[1:66,]
Healthy_matrixall <- zscorematrix[67:158,]
Polyp_matrixall <- zscorematrix[159:234,]

H1 <- Heatmap(scale(t(CRC_matrixall)), column_title = "CRC", row_title = "Metabolites",cluster_rows = FALSE, clustering_method_rows = "complete", show_row_names = FALSE, show_column_names = FALSE, show_column_dend = FALSE, show_heatmap_legend = FALSE) 

H2 <- Heatmap(scale(t(Healthy_matrixall)), column_title = "Healthy", cluster_rows = FALSE, clustering_method_rows = "complete", show_row_names = FALSE, show_column_names = FALSE, show_column_dend = FALSE, show_heatmap_legend = FALSE) 

H3 <- Heatmap(scale(t(Polyp_matrixall)), column_title = "Polyp",cluster_rows = FALSE, clustering_method_rows = "complete", show_row_names = FALSE, show_column_names = FALSE, show_column_dend = FALSE, name = 'Z-score') 

H1+H2+H3

```


***

## Statistical Analysis

Checked data distribution using histograms

Data is not normally distributed so we need to use a non parametric test 
```{r}
#Check data distribution 
par(mfrow=c(2,2))
hist(colon_test_excel$`Glyceraldehyde (89.0 / 59.0)`, main="Histogram (Glyceraldehyde)", col = 'steelblue')
hist(colon_test_excel$`Glucose (179.0 / 89.0)`, main="Histogram (Glucose)", col = 'steelblue')
hist(colon_test_excel$`Pyruvate (87.0 / 43.0)`, main="Histogram (Pyruvate)", col = 'steelblue')
hist(colon_test_excel$`lactate (89.0 / 43.0)`, main="Histogram (Lactate)", col = 'steelblue')
```



***

## Statistical Analysis 

Since I have 3 groups I thought I should do a Kruskal-Wallis test with a pairwise comparison

However, on the paper they did Mann Whitney U tests which only looks at 2 groups at a time 

I tried both tests and I noticed that I got significant p-values between the same groups even though the p-value was not always the same

I decided to move forward with the Mann Whitney test because I care about differences between the pairs of groups


```{r, include=FALSE}
#Kruskal-Walis between 3 groups
#I wanted to try both tests to see if there would be a difference
kruskal.test(`Glyceraldehyde (89.0 / 59.0)` ~ Groups, data = colon_test_excel)

pairwise.wilcox.test(colon_test_excel$`Glyceraldehyde (89.0 / 59.0)`, colon_test_excel$Groups, p.adjust.method = "BH")

```

***

## Statistical Analysis (Mann-Whitney U Test)

Compare CRC and Healthy
```{r}
#Compare CRC and healthy
wilcox.test(crc_healthy$`Glyceraldehyde (89.0 / 59.0)` ~ crc_healthy$Groups, paired = FALSE) 

```

Compare Healthy and Polyp
```{r}
wilcox.test(healthy_polyp$`Glyceraldehyde (89.0 / 59.0)` ~ healthy_polyp$Groups, paired = FALSE) 
```

Compare CRC and Polyp
```{r}
wilcox.test(crc_polyp$`Glyceraldehyde (89.0 / 59.0)` ~ crc_polyp$Groups, paired = FALSE) 
```


***

## Statistical Analysis (Mann-Whitney U Test)

Compare CRC and Healthy
```{r}
#Compare CRC and healthy
wilcox.test(crc_healthy$`Glucose (179.0 / 89.0)` ~ crc_healthy$Groups, paired = FALSE) 
```

Compare Healthy and Polyp
```{r}
wilcox.test(healthy_polyp$`Glucose (179.0 / 89.0)` ~ healthy_polyp$Groups, paired = FALSE) 
```

Compare CRC and Polyp
```{r}
wilcox.test(crc_polyp$`Glucose (179.0 / 89.0)` ~ crc_polyp$Groups, paired = FALSE) 
```


***

## Statistical Analysis (Mann-Whitney U Test)

Compare CRC and Healthy
```{r}
#Compare CRC and healthy
wilcox.test(crc_healthy$`Pyruvate (87.0 / 43.0)` ~ crc_healthy$Groups, paired = FALSE) 
```

Compare Healthy and Polyp
```{r}
wilcox.test(healthy_polyp$`Pyruvate (87.0 / 43.0)` ~ healthy_polyp$Groups, paired = FALSE) 
```

Compare CRC and Polyp
```{r}
wilcox.test(crc_polyp$`Pyruvate (87.0 / 43.0)` ~ crc_polyp$Groups, paired = FALSE) 
```

***

## Statistical Analysis (Mann-Whitney U Test)

Compare CRC and Healthy
```{r}
#Compare CRC and healthy
wilcox.test(crc_healthy$`lactate (89.0 / 43.0)` ~ crc_healthy$Groups, paired = FALSE) 
```

Compare Healthy and Polyp
```{r}
wilcox.test(healthy_polyp$`lactate (89.0 / 43.0)` ~ healthy_polyp$Groups, paired = FALSE) 
```


Compare CRC and Polyp
```{r}
wilcox.test(crc_polyp$`lactate (89.0 / 43.0)` ~ crc_polyp$Groups, paired = FALSE) 
```


## Summary of Statistical Analysis

Glyceraldehyde: Statistically significant difference between CRC/Healthy and CRC/Polyp

Glucose: Statistically significant difference between CRC/Healthy 

Pyruvate: Statistically significant difference between CRC/Healthy 

Lactate: Statistically significant difference between CRC/Healthy 


***

## Data Visualization

Plot mean to visualize difference

```{r}
#Bar graph
bar1 <- ggplot(colon_test_excel, aes(x = Groups, y = `Glyceraldehyde (89.0 / 59.0)`, fill=Groups)) + stat_summary(fun = mean, geom="bar", color = "black") + stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = .5)+ ggtitle("Glyceraldehyde") + ylab("Mean") +   theme(plot.title = element_text(hjust = 0.5)) + geom_signif(comparisons = list(c("CRC", "Healthy")), annotations="***", y_position = 98000, tip_length = 0.02) + geom_signif(comparisons = list(c("CRC", "Polyp")), annotations="***", y_position = 110000, tip_length = 0.02) 

bar2 <-ggplot(colon_test_excel, aes(x = Groups, y = `Glucose (179.0 / 89.0)`, fill=Groups)) + stat_summary(fun = mean, geom="bar", color = "black") + stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = .5)+ ggtitle("Glucose") + ylab("Mean") + theme(plot.title = element_text(hjust = 0.5)) + geom_signif(comparisons = list(c("CRC", "Healthy")), annotations="*", y_position = 8e+07, tip_length = 0.02) 

bar3 <-ggplot(colon_test_excel, aes(x = Groups, y = `Pyruvate (87.0 / 43.0)`, fill=Groups)) + stat_summary(fun = mean, geom="bar", color = "black") + stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = .5)+ ggtitle("Pyruvate") + ylab("Mean") + theme(plot.title = element_text(hjust = 0.5)) + geom_signif(comparisons = list(c("CRC", "Healthy")), annotations="*", y_position = 240000, tip_length = 0.02)

bar4 <-ggplot(colon_test_excel, aes(x = Groups, y = `lactate (89.0 / 43.0)`, fill=Groups)) + stat_summary(fun = mean, geom="bar", color = "black") + stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = .5)+ ggtitle("Lactate") + ylab("Mean") +   theme(plot.title = element_text(hjust = 0.5)) + geom_signif(comparisons = list(c("CRC", "Healthy")), annotations="*", y_position = 4.2e+07, tip_length = 0.02)

#grid.arrange(bar1, bar2, bar3, bar4, ncol=2)

bar1
bar2
bar3
bar4
```


***

## Logistic Regression

Prepared 2 models for the CRC and Healthy groups

Fit_model_1: glucose as predictor variable

Fit_model_2: glucose, lactate and pyruvate as predictor variables

Comparison by AIC shows that Fit2_model_2 is a worse model

cooks.distance command does not show that there are individual outliers that largely influence our data

```{r}
#Model 1
fit_model_1 <- glm(Groups ~ `Glucose (179.0 / 89.0)`, data=crc_healthy, family="binomial")  

summary(fit_model_1)

oddsRatio <- exp(fit_model_1$coefficients)
oddsRatio

exp(confint(fit_model_1))

```
***

## Logistic Regression
```{r}
#Model 2
fit_model_2 <- glm(Groups ~ `Glucose (179.0 / 89.0)` + `lactate (89.0 / 43.0)` +`Pyruvate (87.0 / 43.0)`, data=crc_healthy, family="binomial")

summary(fit_model_2)

#Comparison using the Akaike Information Criteria
AIC(fit_model_1, fit_model_2)
 
which(cooks.distance(fit_model_1)>10)
which(cooks.distance(fit_model_2)>10)

```


***

## Summary and Future Direction


The dataframe contained a lot of information so I had to focus on a subset

Higher mean levels of glucose, pyruvate and lactate were observed in CRC patients compared to healthy patients

Interestingly, a regression model with glucose, lactate and pyruvate behaved worse than a model containing only glucose

For the future, I would like to analyze a bigger portion of the data and do PLS Discriminant Analysis to identify metabolites that can distinguish between healthy and CRC patients

I would also like to explore if there is a relationship between the different cancer stages and different levels of metabolites


***

## Citations

Zhu, J., et al. (2014). "Colorectal Cancer Detection Using Targeted Serum Metabolic Profiling." Journal of Proteome Research 13(9): 4120-4130.

## THANK YOU




