---
title: "Masters Homework 3"
author: "Fred LaPolla"
date: "7/29/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use a for loop to find mean and standard deviation of A1C, Lead, HDL, GlucoseSI, Creatine, Glucose and LDL Estimate. Print the names as well. 

```{r}

```

Create a vector to put the results of the previous loop into to store them

```{r}

```

Use ifelse to create a three level variable for normal, borderline and high ldl (< 130 = normla, 130-159 = borderline, and >160 = high per medline.gov)

```{r}

```

Use by() to find the average cholsterol level by hsq_1 status:

```{r}

```



>- Subset to only have females (gender == "Female")

```{r}

```




>- Create a subset of males and find the median total cholesterol level

```{r}

```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

```

Find the class of the first three columns of data. What are they?

```{r}

```

How does R treat blank values?

What command can be used to identify where NAs are located in a column?

```{r}

```


How can you use indexing to create a subset without any NAs?

```{r}

```

How could you delete a column of data, for example, the 5th 
```{r}

```


What command would let you find a set of text in the cells of a column?

```{r}

```


Rename the HSQ1 variable to Health_Status

```{r}

```

Explore the Mutate command and create a ratio of total cholesterol divided by HDL

```{r}

```


