---
title: "Masters Homework 4"
author: "Fred LaPolla"
date: "8/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use the Synthetic Mass Data:

1. Make a subset of Patient data to have patient ID, sex/gender, race and ethnicity. 

```{r}

```

2. Make a subset of observation data for body weight and body height, and filter to have only the most recent of each. 

```{r}

```

3. Convert the body weight and height tables to wide format. 

```{r}

```

4. Merge the body weight and height information onto your demographic dataframe created in question one. 

```{r}

```

5. Calculate BMI into a new column. The equation is weight (kg)/height ^2 (m)

```{r}

```




6. Why would one use a visualization for exploratory analysis?



7. What information about the nature of our data can a histogram tell us?



8. Make a histogram on LEAD from the nyc data frame. Is it skewed? Which way? 

```{r}

```

>- Make a QQPlot to assess the normality UCREATININE. Is the distribution normal? How can you tell?

```{r}

```

>- Make a boxplot glucose levels by diabetes diagnosis:

```{r}


```

>- Make a scatterplot of total cholesterol by LDL. Does there appear to be a relationship?

```{r}

```


